---
title: "Bike Share Project"
author: "Tim Fuller"
date: "2/19/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

     
You will answer your questions of interest about the Motivate datasets using R. To help guide your work in this project, a template with helper code and comments is provided in a notebook file on the next page. You will need the three city dataset files too:
editor_options: 
  chunk_output_type: inline
---

Explore Bike Share Data
For this project, your goal is to ask and answer three questions about the available bikeshare data from Washington, Chicago, and New York. This notebook can be submitted directly through the workspace when you are confident in your results.

You will be graded against the project Rubric by a mentor after you have submitted. To get you started, you can use the template below, but feel free to be creative in your solutions!


Question 1
How does each city Compare to the other on a month by month basis?
```{r}
#Question 1

#Initiate Packages
library(ggplot2)
library(ggthemes)
library(lubridate)

#Set Theme
theme_set(theme_minimal(8))

#Data Import
ny <- read.csv('new-york-city.csv')
wash <- read.csv('washington.csv')
chi <- read.csv('chicago.csv')

#Append City Name to Columns
ny$city <- 'New York'
wash$city <- 'Washington'
chi$city <- 'Chicago'

#Removes Gender and Birth.Year to match data with Washington
matchCitySet = function(cty)
{
  cty$Gender <- NULL
  cty$Birth.Year <- NULL
  return(cty)
}

#Matching Data to Washington
ny <- matchCitySet(ny)
chi <- matchCitySet(chi)

#Combine Cities to full set
fullSet <-  rbind(ny,wash,chi)

#Count By Month Per City
ggplot(aes(x=city), data=fullSet) +
  geom_histogram(stat='count') +
  ggtitle("Histogram of Bike Share Count Per Month by City") +
  labs(x="City", y="Trip Count") +
  scale_y_continuous(limits = c(0,100000), breaks = seq(0,100000,10000), labels=scales::comma) +
  facet_wrap(~month(Start.Time,TRUE), scale="free")

#The Below Summary Shows the June Spike
by(month(fullSet$Start.Time, TRUE),fullSet$city,summary)
```

Here we see each city getting a spike in users in June, with Chicago seeing the largest.  
It would be interested to map this data with relevant weather information.


Question 2
What is the Most Popular Day Of Bike Rentals?

```{r}
#Initiate Packages
library(ggplot2)
library(ggthemes)
library(lubridate)

#Set Theme
theme_set(theme_minimal(12))

#Data Import
ny <- read.csv('dataSets\\new-york-city.csv')
wash <- read.csv('dataSets\\washington.csv')
chi <- read.csv('dataSets\\chicago.csv')

#Append City Name to Columns
ny$city <- 'New York'
wash$city <- 'Washington'
chi$city <- 'Chicago'

#Removes Gender and Birth.Year to match data with Washington
matchCitySet = function(cty)
{
  cty$Gender <- NULL
  cty$Birth.Year <- NULL
  return(cty)
}

#Matching Data to Washington
ny <- matchCitySet(ny)
chi <- matchCitySet(chi)

#Combine Cities to full set
fullSet <-  rbind(ny,wash,chi)

#Ordered Factor To Keep Weekdays in order
fullSet$day <- factor(weekdays(as.Date(fullSet$Start.Time, abbreviate=TRUE)),
                      levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

ggplot(fullSet, aes(x=fullSet$day, y=stat(count), fill=fullSet$city)) +
  geom_bar(position='dodge') +
  labs(x="Day of Week", y="Rental Count", fill = "City") +
  ggtitle("Bikeshare Rental Count by Day of Week", subtitle = "New York Bikers Love Wednesday") +
  scale_y_continuous(limits = c(0,55000), breaks = seq(0,55000,5000),
                     labels=scales::comma)

#You Can see the Wednesday Spike Below
by(fullSet$day,fullSet$city,summary)
```



Here we see slight elevations during weekdays.  Most likely bikers commuting to work.  
Wednesday is a bit of an outlier for New York and Washington. 

Question 3
How do trip durations compare across the 3 cities?
```{r}
#Initiate Packages
library(ggplot2)
library(ggthemes)

#Set Theme
theme_set(theme_minimal(12))

#Data Import
ny <- read.csv('dataSets\\new-york-city.csv')
wash <- read.csv('dataSets\\washington.csv')
chi <- read.csv('dataSets\\chicago.csv')

#Removes Gender and Birth.Year to match data with Washington
matchCitySet = function(cty)
{
  cty$Gender <- NULL
  cty$Birth.Year <- NULL
  return(cty)
}

#Matching Data to Washington
ny <- matchCitySet(ny)
chi <- matchCitySet(chi)

#Append City Name to Columns
ny$city <- 'New York'
wash$city <- 'Washington'
chi$city <- 'Chicago'

#Combine Cities to full set
fullSet <-  rbind(ny,wash,chi)

##Trip Duration Analysis By City
ggplot(aes(x=city,y=Trip.Duration/60), data=na.omit(fullSet), group=1) +
  geom_point(
    alpha=1/10, 
    position = position_jitter(h=0),
    color='grey30') +
  scale_y_continuous(limits = c(0,480), breaks = seq(0,480,60)) +
  labs(x="City", y="Trip Duration (Minutes)") +
  ggtitle("Trip Duration By City", subtitle = "Washington generally has longer durations but maintains a similar average time.") +
  stat_summary(fun.y=mean, aes(group=1), geom="line",linetype=2, color="blue")

by(fullSet$Trip.Duration/60,fullSet$city,summary)

#The below formula shows the mean trip duration in minutes
#You can see that washington has about a 5 minute higher average
by(fullSet$Trip.Duration/60,fullSet$city,mean)
  
```

Here we can see that Washington trends towards longer trips over the other two cities.
We can also see the data has an exceptionally long tail which is why the limits are set
to 8 hour trip durations for the evaluation. 